---
output: github_document
bibliography: "references.bib"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


knitr::knit_hooks$set(margin = function(before, options, envir){
  if (before){
    par(mar=c(0.1,0.1,0.1,0.1))
  } 
})

```

# potential

This package provides functions to compute potential models as
defined by @STEWART89. Several options
are available to customize the models, such as the possibility to fine-tune
the distance friction functions or to use custom distance matrices. Some
computations are parallelized to improve their efficiency.


## Installation

### From GitHub
Development version / unstable

```{r, eval = FALSE}
require(remotes)
remotes::install_github("riatelab/potential")
```

## Demo

```{r, margin = TRUE}
library(sf)
library(potential)

plot(st_geometry(n3_poly), col = "lightblue")
plot(st_geometry(n3_pt), add = TRUE)


g <- create_grid(x = n3_poly, res = 100000)

plot(st_geometry(n3_poly), col = "lightblue")
plot(st_geometry(n3_pt), add = TRUE)
plot(st_geometry(g), cex = .2, add = TRUE)


d <- create_matrix(x = n3_pt, g)
knitr::kable(d[1:5, 1:5], row.names = T)


```

```{r}
prob_interaction(fun = "e", span = 75000, beta = 2, limit = 250000)
```

```{r, margin=TRUE}
pot <- potential(
  x = n3_pt, y = g,
  d = d, var = "POP19",
  fun = "e", span = 75000,
  beta = 2
)
plot(pot["OUTPUT"], reset = FALSE, pch= 20, cex = 1.2)
plot(st_geometry(n3_poly), add = TRUE)

```



```{r, margin = TRUE}
equipot <- equipotential(pot, mask = n3_poly)
plot(st_geometry(equipot), col = hcl.colors(nrow(equipot), "cividis"))
```



## Note

This package provides access to the revamped potential-related functions 
initialy offered by 
[`SpatialPosition`](https://CRAN.R-project.org/package=SpatialPosition).

## References









