---
output: github_document
bibliography: "vignettes/references.bib"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "man/figures/")
```



# potential

[![R-CMD-check](https://github.com/riatelab/potential/workflows/R-CMD-check/badge.svg)](https://github.com/riatelab/potential/actions)


This package provides functions to compute potential models as
defined by @STEWART41. Several options
are available to customize the models, such as the possibility to fine-tune
the distance friction functions or to use custom distance matrices. Some
computations are parallelized to improve their efficiency.


## Installation

### From GitHub
Development version / unstable

```{r, eval = FALSE}
require(remotes)
remotes::install_github("riatelab/potential")
```

## Demo


```{r, fig.width = 7, fig.height= 7}
library(sf)
library(potential)
library(cartography)
# create a regular grid
y <- create_grid(x = n3_poly, res = 20000)

# compute potentials
pot <- mcpotential(
  x = n3_pt, y = y,
  var = "POP19",
  fun = "e", span = 75000,
  beta = 2, limit = 250000
)
y$pot <- pot / max(pot) * 100

# create equipotential areas
equipot <- equipotential(y, var = "pot", mask = n3_poly)

# map potentials
par(mar = c(0, 0, 1.2, 0), bg = "#b5bece")
choroLayer(equipot, var = "center", breaks = seq(0,100,length.out = 11), 
           col = hcl.colors(10, 'teal'),
           border = "#121725", legend.pos = "bottom", 
           lwd = .2, legend.title.txt = "Potential Intensity",
           legend.horiz = T)
layoutLayer(title = "Potentials of Population", 
            col = "#121725", coltitle = "#4dB8da",
            sources = "© EuroGeographics for the administrative boundaries and © Eurostat for data",
            horiz = F, postitle = "center", scale = F)
```



## Note

This package provides access to the revamped potential-related functions 
initialy offered by 
[`SpatialPosition`](https://CRAN.R-project.org/package=SpatialPosition).

## References









