---
title: "Stewart Potentials: a Use Case"
output: rmarkdown::html_vignette
bibliography: "references.bib"
vignette: >
  %\VignetteIndexEntry{Stewart Potentials: a Use Case}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE,
  collapse = TRUE,
  comment = "#>", 
  fig.width = 7,
  fig.height = 6
)


```


The Stewart model is a spatial interaction modeling approach which aims to compute indicators based on stock values weighted by distance. These indicators have two main interests:

1. they produce understandable maps by smoothing complex spatial patterns;
2. they enrich stock variables with contextual spatial information.

At the European scale, this functional semantic simplification may help to show a smoothed context-aware picture of the localized socio-economic activities.

In this vignette, we show a use case of these potentials on the regional GDP per capita at the European scale with three maps:

* a regional map of the GDP per capita;
* a regional map of the potential GDP per capita;
* a smoothed map of the GDP per capita.

Note that this example is based on data proposed in the `[cartography]( https://cran.r-project.org/package=cartography)` package but function used to create maps are from `[mapsf]( https://cran.r-project.org/package=mapsf)`.  

## Regional Map of the GDP per Capita

```{r regionalmap, fig.width=7, fig.height=6, fig.show='hold' }
library(cartography)
library(mapsf)
library(potential)
library(sp)
data(nuts2006)
# Transform sp data to sf data
nuts3.spdf@data <- nuts3.df
nuts3 <- st_as_sf(nuts3.spdf)
nuts0 <- st_as_sf(nuts0.spdf)
world <- st_as_sf(world.spdf)

# Compute the GDP per capita variable
nuts3$gdpcap <- nuts3$gdppps2008 * 1000000 / nuts3$pop2008
# Discretize the variable
bv <- quantile(nuts3$gdpcap, seq(from = 0, to = 1, length.out = 9))
# Draw the map
# Set a color palette
pal <- mf_get_pal(n = 9, palette = "Burg", rev = TRUE)
# Draw the basemap
mf_theme(bg = "#cdd2d4", mar = c(0,0,1.2,0), tab = FALSE)
mf_init(nuts0)
mf_map(world, col = "#f5f5f3ff", border = "#a9b3b4ff", add = TRUE)
# Map the regional GDP per capita
mf_map(x = nuts3, var = "gdpcap", type = "choro",
       leg_pos = "topright",
       breaks = bv, 
       pal = pal, 
       border = NA, 
       leg_title = "GDP per capita",
       leg_val_rnd = -2, 
       add = TRUE)
mf_map(nuts0, add = TRUE, lwd = 0.5, border = "grey30", col = NA)
mf_map(world, col = NA, border = "#7DA9B8", add = TRUE)
# Set a layout
mf_title(txt = "GDP per capita Inequality in Europe")
mf_credits(txt = "Basemap: UMS RIATE, 2015 - Data: Eurostat, 2008\nT. Giraud, 2020", 
           bg = "#ffffff80")
```


## Regional Map of the Potential GDP per Capita

We compute the potentials of GDP for each spatial unit. The computed value takes into account the spatial distribution of the stock variable and return a sum weighted by distance, according a specific spatial interaction and fully customizable function.

```{r regionalmappot, fig.width=7, fig.height=6, fig.show='hold' }

# Create a distance matrix between units
nuts3pt <- nuts3
st_geometry(nuts3pt) <- st_centroid(st_geometry(nuts3pt))
d <- create_matrix(nuts3pt, nuts3pt)

# Compute the potentials of population and GDP per units
# function = exponential, beta = 2, span = 75 km
pot <- potential(x = nuts3pt, 
                 y = nuts3pt, 
                 d = d, 
                 var = c("pop2008", "gdppps2008"), 
                 fun = "e",  
                 beta = 2, 
                 span = 75000)
# A the potential GDP per capita
nuts3$gdpcappot <-  pot[, 2] * 1000000 / pot[, 1]

# Discretize the variable
bv2 <- c(min(nuts3$gdpcappot), bv[2:8], max(nuts3$gdpcappot))
# Draw the map
# Draw the basemap
mf_init(nuts0)
mf_map(world, col = "#f5f5f3ff", border = "#a9b3b4ff", add = TRUE)
# Map the regional potential of GDP per capita
mf_map(x = nuts3, var = "gdpcappot", type = "choro",
       leg_pos = "topright",
       breaks = bv2, 
       pal = pal, 
       border = NA, 
       leg_title = "Potential\nGDP per capita",
       leg_val_rnd = -2, 
       add = TRUE)
mf_map(nuts0, add = TRUE, lwd = 0.5, border = "grey30", col = NA)
mf_map(world, col = NA, border = "#7DA9B8", add = TRUE)
# Set a text to explicit the function parameters
text(x = 6271272, y = 3743765, 
     labels = "Distance function:\n- type = exponential\n- beta = 2\n- span = 75 km", 
     cex = 0.8, adj = 0, font = 3)
# Set a layout
mf_title(txt = "GDP per capita Inequality in Europe")
mf_credits(txt = "Basemap: UMS RIATE, 2015 - Data: Eurostat, 2008\nT. Giraud, 2020", 
           bg = "#ffffff80")
```

This map gives a smoothed picture of the spatial patterns of wealth in Europe while keeping the original spatial units as interpretive framework. Hence, the map reader can still rely on a known territorial division to develop its analyses.


## Smoothed Map of the GDP per Capita

In this case, the potential GDP per capita is computed on a regular grid.

```{r smoothedmappot, fig.width=7, fig.height=6, fig.show='hold' }
# Compute the potentials of population on a regular grid (50km span)
g <- create_grid(x = nuts3, res = 50000)
d <- create_matrix(nuts3pt, g)
# function = exponential, beta = 2, span = 75 km
pot2 <- potential(x = nuts3pt, 
                  y = g, 
                  d = d, 
                  var = c("pop2008", "gdppps2008"), 
                  fun = "e",  
                  beta = 2, 
                  span = 75000)
# Create the ratio variable
g$gdpcappot <-  pot2[, 2] * 1000000 / pot2[, 1]

# Create an isopleth layer
pot <- equipotential(x = g, var = "gdpcappot", breaks = bv, mask = nuts3)
# Get breaks values
bv3 <- sort(c(unique(pot$min), max(pot$max)), decreasing = FALSE)
# Draw the map
# Draw the basemap
mf_init(nuts0)
mf_map(world, col = "#f5f5f3ff", border = "#a9b3b4ff", add = TRUE)
# Map the potential GDP per Capita
mf_map(x = pot, var = "center", type = "choro",
       leg_pos = "topright",
       breaks = bv3, 
       pal = pal, 
       border = NA, 
       leg_title = "Potential\nGDP per capita",
       leg_val_rnd = -2, 
       add = TRUE)
mf_map(nuts0, add = TRUE, lwd = 0.5, border = "grey30", col = NA)
mf_map(world, col = NA, border = "#7DA9B8", add = TRUE)
# Set a text to explicit the function parameters
text(x = 6271272, y = 3743765, 
     labels = "Distance function:\n- type = exponential\n- beta = 2\n- span = 75 km", 
     cex = 0.8, adj = 0, font = 3)
# Set a layout
mf_title(txt = "GDP per capita Inequality in Europe")
mf_credits(txt = "Basemap: UMS RIATE, 2015 - Data: Eurostat, 2008\nT. Giraud, 2020", 
           bg = "#ffffff80")
```

Unlike the previous maps, this one doesn't keep the initial territorial division to give a smoothed picture of the spatial patterns of wealth in Europe. The result is easy to read and can be considered as a bypassing of the Modifiable Areal Unit Problem (MAUP).
